---
title: "Project"
author: "Rafael Asuncion and Erik Babb"
date: "May 18, 2017"
output: word_document
---
```{r enable parallel processing}
library(doParallel)
registerDoParallel(makeCluster(detectCores()))
```

Step1: Collecting the data

For this data exploration, we shall be examining the Ames Housing Dataset
```{r Load data}
#housing <- read.csv("C:/Users/Rafael Asuncion/Documents/WORK/Capstone/train.csv")
housing <- read.csv("C:/Users/eb832f/Documents/Data Science/CSUEB/Spring 2017/Capstone/Raf/train.csv")
#train <- read.csv("C:/Users/eb832f/Documents/Data Science/CSUEB/Spring 2017/Capstone/train.csv")
```

Step2: Examining the data

-observe summary of data. value to be predicted is salesPrice.
-drop id as an independent variable
```{r }
housing <- housing[ , !(names(housing) %in% "Id")] #remove the Id column
str(housing)
```

-examine distribution of sales price
```{r }
summary(housing$SalePrice)
hist(housing$SalePrice, xlab = "Sale Prices", main = "Histogram of Sale Price", col = "green")

```

-examine distribution of property ages
```{r }
hist(housing$YearBuilt, xlab = "Year Built", main = "Histogram of Year Built", col = "blue")

```

-examine LotArea and 1st Floor Square Feet
```{r}
library(ggplot2)
ggplot(data=housing, aes(x=LotArea, y=X1stFlrSF)) + geom_point(color="black")
ggplot(data=housing, aes(x=LotArea, y=X1stFlrSF)) + geom_point(aes(color=SalePrice)) +
  xlim(0, 50000) + 
  ylim(0, 4000)
```

- examine correlations of various independent variables & dependent variable SalePrice.
```{r }
cor(housing[c("SalePrice", "LotArea", "OverallQual", "YearBuilt", "YrSold", "OverallCond" , "GrLivArea", "PoolArea")])

```

-scatterplot matrix for variables

```{r }
pairs(housing[c("SalePrice", "LotArea", "OverallQual", "YearBuilt", "YrSold", "OverallCond" , "GrLivArea", "PoolArea")])

```

- another view of scatterplot for variables
```{r }
library(psych)
pairs.panels(housing[c("SalePrice", "LotArea", "OverallQual", "YearBuilt", "YrSold", "OverallCond" , "GrLivArea", "PoolArea")])
```

- examine missing values
```{r }
#View all the NAs
na_count <- sapply(housing, function(y) sum(length(which(is.na(y)))))
na_count <- data.frame(na_count)
na_count

library(Amelia)
missmap(housing, main = "Missing values vs observed")
```

- include NA as a factor for categorical variables with NA (not applicable) as an option
```{r }
housing$Alley <- addNA(housing$Alley)
housing$PoolQC <- addNA(housing$PoolQC)
housing$MiscFeature <- addNA(housing$MiscFeature)
housing$Fence <- addNA(housing$Fence)
housing$FireplaceQu <- addNA(housing$FireplaceQu)
housing$GarageCond <- addNA(housing$GarageCond)
housing$GarageQual <- addNA(housing$GarageQual)
housing$GarageFinish <- addNA(housing$GarageFinish)
housing$GarageType <- addNA(housing$GarageType)
housing$BsmtQual <- addNA(housing$BsmtQual)
housing$BsmtCond <- addNA(housing$BsmtCond)
housing$BsmtExposure <- addNA(housing$BsmtExposure)
housing$BsmtFinType1 <- addNA(housing$BsmtFinType1)
housing$BsmtFinType2 <- addNA(housing$BsmtFinType2)
```

- examine missing values again
```{r }
#View all the NAs
na_count <- sapply(housing, function(y) sum(length(which(is.na(y)))))
na_count <- data.frame(na_count)
na_count

library(Amelia)
missmap(housing, main = "Missing values vs observed")
```

- Set 1
- drop LotFrontage, GarageYrBlt, MasVnrArea, MasVnrType, Electrical because of missing values wherein NA not as an option.
```{r }
train_DropNAs <- housing[ , !(names(housing) %in% c("LotFrontage","GarageYrBlt","MasVnrArea","MasVnrType", "Electrical"))]
write.csv(train_DropNAs, file = "train_DropNAs.csv",row.names=FALSE)
```

- Set 2
- Convert NAs to mode in LotFrontage, GarageYrBlt, MasVnrArea, MasVnrType, Electrical instead
```{r }
train_NAsToMode <- housing

mode <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

train_NAsToMode["LotFrontage"][is.na(train_NAsToMode["LotFrontage"])] <- mode(housing$LotFrontage)
train_NAsToMode["GarageYrBlt"][is.na(train_NAsToMode["GarageYrBlt"])] <- mode(housing$GarageYrBlt)
train_NAsToMode["MasVnrArea"][is.na(train_NAsToMode["MasVnrArea"])] <- mode(housing$MasVnrArea)
train_NAsToMode["MasVnrType"][is.na(train_NAsToMode["MasVnrType"])] <- mode(housing$MasVnrType)
train_NAsToMode["Electrical"][is.na(train_NAsToMode["Electrical"])] <- mode(housing$Electrical)
write.csv(train_NAsToMode, file = "train_NAsToMode.csv",row.names=FALSE)
```

- Set 3
- convert features with data type integer to factor
```{r}
#Convert Integer to Factor
train_IntsToFactors <- housing
train_IntsToFactors$MSSubClass <- as.factor(train$MSSubClass)
train_IntsToFactors$KitchenAbvGr <- as.factor(train$KitchenAbvGr)
train_IntsToFactors$OverallCond <- as.factor(train$OverallCond)
train_IntsToFactors$YrSold <- as.factor(train$YrSold)
train_IntsToFactors$MoSold <- as.factor(train$MoSold)
write.csv(train_IntsToFactors, file = "train_IntsToFactors.csv",row.names=FALSE)
```


- Set 4
- Remove all nominal variables
```{r}

nomvars <- c("MSZoning", "Street", "Alley", "LandContour", "LotConfig",
             "Neighborhood", "Condition1", "Condition2", "HouseStyle",
             "RoofStyle", "RoofMatl", "Exterior1st", "Exterior2nd",
             "MasVnrType", "Foundation", "Heating", "CentralAir", "Electrical",
             "GarageType", "PavedDrive", "MiscFeature", "SaleType",
             "SaleCondition", "MSSubClass", "LotShape", "Utilities", "LandSlope", "BldgType",
             "OverallQual", "OverallCond", "ExterQual", "ExterCond", "BsmtQual",
             "BsmtCond", "BsmtExposure", "BsmtFinType1", "BsmtFinType2",
             "HeatingQC", "KitchenQual", "Functional", "FireplaceQu",
             "GarageFinish", "GarageQual", "GarageCond", "PoolQC", "Fence",
             "MoSold"
             )
train_NoNomVars <- housing[ , !(names(housing) %in% nomvars)]
write.csv(train_NoNomVars, file = "train_NoNomVars.csv",row.names=FALSE)
```


- Set 5
- Explicitly define ordinality for applicable factor features
```{r}
train_OrdDefined <- housing

#covert NAs to more meaningful name
levels(train_OrdDefined$BsmtQual) <- c(levels(train_OrdDefined$BsmtQual), "noBasement")
levels(train_OrdDefined$BsmtCond) <- c(levels(train_OrdDefined$BsmtCond), "noBasement")
levels(train_OrdDefined$BsmtExposure) <- c(levels(train_OrdDefined$BsmtExposure), "noBasement")
levels(train_OrdDefined$BsmtFinType1) <- c(levels(train_OrdDefined$BsmtFinType1), "noBasement")
levels(train_OrdDefined$BsmtFinType2) <- c(levels(train_OrdDefined$BsmtFinType2), "noBasement")
levels(train_OrdDefined$FireplaceQu) <- c(levels(train_OrdDefined$FireplaceQu), "noFireplace")
levels(train_OrdDefined$GarageFinish) <- c(levels(train_OrdDefined$GarageFinish), "noGarage")
levels(train_OrdDefined$GarageQual) <- c(levels(train_OrdDefined$GarageQual), "noGarage")
levels(train_OrdDefined$GarageCond) <- c(levels(train_OrdDefined$GarageCond), "noGarage")
levels(train_OrdDefined$PoolQC) <- c(levels(train_OrdDefined$PoolQC), "noPool")
levels(train_OrdDefined$Fence) <- c(levels(train_OrdDefined$Fence), "noFence")

lvlOrd <- list(
  LotShape = c("Reg", "IR1", "IR2", "IR3"),
  Utilities = c("AllPub", "NoSewr", "NoSeWa", "ELO"),
  LandSlope = c("Gtl", "Mod", "Sev"),
  BldgType = c("1Fam", "2FmCon", "Duplx", "TwnhsE", "TwnhsI"),
  OverallQual = 10:1,
  OverallCond = 10:1,
  ExterQual = c("Ex", "Gd", "TA", "Fa", "Po"),
  ExterCond = c("Ex", "Gd", "TA", "Fa", "Po"),
  BsmtQual = c("Ex", "Gd", "TA", "Fa", "Po", "noBasement"),
  BsmtCond = c("Ex", "Gd", "TA", "Fa", "Po", "noBasement"),
  BsmtExposure = c("Gd", "Av", "Mn", "No", "noBasement"),
  BsmtFinType1 = c("GLQ", "ALQ", "BLQ", "Rec", "LwQ", "Unf", "noBasement"),
  BsmtFinType2 = c("GLQ", "ALQ", "BLQ", "Rec", "LwQ", "Unf", "noBasement"),
  HeatingQC = c("Ex", "Gd", "TA", "Fa", "Po"),
  KitchenQual = c("Ex", "Gd", "TA", "Fa", "Po"),
  Functional = c("Typ", "Min1", "Min2", "Mod", "Maj1", "Maj2", "Sev", "Sal"),
  FireplaceQu = c("Ex", "Gd", "TA", "Fa", "Po", "noFireplace"),
  GarageFinish = c("Fin", "RFn", "Unf", "noGarage"),
  GarageQual = c("Ex", "Gd", "TA", "Fa", "Po", "noGarage"),
  GarageCond = c("Ex", "Gd", "TA", "Fa", "Po", "noGarage"),
  PoolQC = c("Ex", "Gd", "TA", "Fa", "noPool"),
  Fence = c("GdPrv", "MnPrv", "GdWo", "MnWw", "noFence"),
  MoSold = 1:12
)
for (j in names(lvlOrd)) {
  train_OrdDefined[[j]] <- factor(train_OrdDefined[[j]], levels = lvlOrd[[j]], ordered = TRUE)
}

write.csv(train_OrdDefined, file = "train_OrdDefined.csv",row.names=FALSE)
```


- splitting data into training, validation and test.
```{r }
set.seed(300)
indx = sample(1:nrow(housing), as.integer(0.60*nrow(housing)))
housing_train = housing[indx,]
housing_test = housing[-indx,]

train1 <- train_DropNAs[indx,]
test1 <- train_DropNAs[-indx,]

train2 <- train_NAsToMode[indx,]
test2 <- train_NAsToMode[-indx,]

train3 <- train_IntsToFactors[indx,]
test3 <- train_IntsToFactors[-indx,]

train4 <- train_NoNomVars[indx,]
test4 <- train_NoNomVars[-indx,]

train5 <- train_OrdDefined[indx,]
test5 <- train_OrdDefined[-indx,]

#What does this do? train and test are already created above
#indx2 = sample(1:nrow(housing_test), as.integer(0.6*nrow(housing_test)))
#housing_validation = housing_test[-indx2,]
#housing_test = housing_test[indx2,]
```


-Create a table to display the the RSME result values
```{r}
mat = matrix(0, nrow = 5, ncol = 5)
dat = as.data.frame(mat)
colnames(dat) <- c("Basic LM", "Adv LM", "Bagging", "rpart", "Caret RF")
rownames(dat) <- c("train_DropNAs", "train_NAsToMode", "train_IntsToFactors", "train_NoNomVars", "train_OrdDefined")

```

Model A: LINEAR Regression Model 
Step3: Training the model
```{r }
model <- lm(SalePrice ~ LotArea + OverallQual + GrLivArea + GarageCars + YearBuilt + TotalBsmtSF + WoodDeckSF, data = housing_train)
summary(model)
```

Step4: Measuring Performance

- make predictions and evaluate RMSE
```{r }
my_predictions <- predict(model, housing_test)

rmse <- function(error)
{
    sqrt(mean(error^2))
}

housing_test_actual = housing_test[, "SalePrice"]
error <- housing_test_actual - my_predictions
rmse(error)
 
```

Step5: Improving Performance

- include more significant predictors. As seen in output, this improves RSquare for training data since more predictors are added.
```{r }
model2 <- lm(SalePrice ~ LotArea + OverallQual + GrLivArea  + YearBuilt + TotalBsmtSF + WoodDeckSF + BedroomAbvGr + KitchenAbvGr + GarageArea  + ScreenPorch , data = housing_train)
summary(model2)
```

- make predictions and evaluate RMSE. 
```{r }
my_predictions2  <-predict(model2, housing_test)

error <- housing_test_actual - my_predictions2
rmse(error)
 
```

ModelB: Regression Tree

Step 3: Training the model

```{r }
library(rpart)
model3_decisiontree1 <- rpart(SalePrice ~ ., data = housing_train)
summary(model3_decisiontree1) 
```

- Step 4: Evaluating Model Performance

- use rpart decision tree and make predictions. examine rmse.
```{r }
my_prediction_decision_tree <- predict(model3_decisiontree1, housing_test)
error_decision_tree <- housing_test_actual - my_prediction_decision_tree
rmse(error_decision_tree)
 
```

- Step 5: Improving Model Performance

- implement a Bagged Regeression Tree

```{r }
library(ipred)
set.seed(300)
model4_decision_tree_ipred <- bagging(SalePrice ~ ., data = housing, nbagg = 25)
```

- make predictions and examine rmse. 

```{r }
my_prediction_decision_tree_ipred <- predict(model4_decision_tree_ipred, housing_test)
error_decision_tree_ipred <- housing_test_actual - my_prediction_decision_tree_ipred
rmse(error_decision_tree_ipred)
```

ModelC: Random Forest

Step3: Training the model

- implement a random forest using significant continous variable predictors
```{r }
library(randomForest)
set.seed(300)
rf <- randomForest(SalePrice ~ LotArea + OverallQual + GrLivArea + GarageCars + YearBuilt + TotalBsmtSF + WoodDeckSF, data = housing_train, na.action = na.omit)
rf
```

Step4 : Evaluating model performance

```{r }
predictions_rf <- predict(rf, housing_test)
error_rf <- housing_test_actual - predictions_rf
rmse(error_rf)
```

Step 5: Improving model performance

- auto-tune random forest
```{r }
library(caret)
#ctrl <- trainControl(method = "repeatedcv", number = 10, repeats = 10) #slower

ctrl <- trainControl(method = "cv", number = 5) #faster

# auto-tune a random forest
grid_rf <- expand.grid(.mtry = c(2, 4, 8, 16))

set.seed(300)
m_rf <- train(SalePrice ~ ., data = housing_train, method = "rf", trControl = ctrl,
              tuneGrid = grid_rf, na.action = na.omit)
m_rf
```

- examine rmse of random forest
```{r }
predictions_mrf <- predict(m_rf, housing_test)
error_mrf <- housing_test_actual - predictions_mrf
rmse(error_mrf)
```

```{r}
m_rf1 <- train(SalePrice ~ ., data = train1, method = "rf", trControl = ctrl,
              tuneGrid = grid_rf, na.action = na.omit)
predictions_mrf1 <- predict(m_rf1, test1)
error_mrf1 <- test1[, "SalePrice"] - predictions_mrf1
rmse(error_mrf1)

```

